---
title: "Dataset Management and Format Conversion"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dataset Management and Format Conversion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## Overview

diversityGPT provides a comprehensive dataset management system that handles:

- **30+ Built-in Datasets**: Curated collection with metadata
- **Format Converters**: BIOM, QIIME2, MetaPhlAn, and more
- **Precomputed Analyses**: Instant access to universal transformations
- **Smart Caching**: Automatic storage of processed results

```{r setup}
library(diversityGPT)
library(phyloseq)
library(DT)
```

## Dataset Discovery

### Browsing Available Datasets

```{r list-datasets}
# List all available datasets
all_datasets <- list_available_datasets()

# View first few entries
head(all_datasets[, c("name", "type", "samples", "taxa", "source")])
```

### Filtering Datasets

Find datasets that match your criteria:

```{r filter-datasets}
# Find 16S datasets with more than 100 samples
large_16s <- list_available_datasets(type = "16S")
large_16s <- large_16s[!is.na(large_16s$samples) & large_16s$samples > 100, ]

if (nrow(large_16s) > 0) {
  print(large_16s[, c("name", "samples", "description")])
} else {
  cat("No 16S datasets with >100 samples found\n")
}
```

### Searching by Keywords

```{r search-datasets}
# Search for gut-related datasets
gut_datasets <- search_datasets("gut")
print(gut_datasets[, c("name", "type", "description")])

# Search for time series data
time_series <- search_datasets("time")
print(time_series[, c("name", "description")])
```

## Loading Datasets

### Quick Load

The simplest way to load a dataset:

```{r quick-load}
# Load by ID
physeq <- load_dataset("enterotype")
print(physeq)
```

### Load with Metadata

Get detailed information before loading:

```{r load-with-info}
# Get dataset information
info <- get_dataset_info("globalpatterns")

# Examine metadata
cat("Dataset:", info$name, "\n")
cat("Samples:", info$samples, "\n")
cat("Taxa:", info$taxa, "\n")
cat("Citation:", info$citation, "\n")

# Load if it matches your needs
if (info$has_tree) {
  gp_data <- load_dataset("globalpatterns")
}
```

### Precomputed Datasets

Load datasets with pre-calculated universal transformations:

```{r precomputed}
# List precomputed datasets
precomputed <- list_available_datasets(source = "precomputed")
print(precomputed[, c("name", "samples", "taxa")])

# Load with precomputed results
result <- load_dataset("globalpatterns_demo", return_universal = TRUE)

# If return_universal = TRUE, you get both phyloseq and universal info
if (is.list(result) && "universal_info" %in% names(result)) {
  physeq <- result$phyloseq
  universal_info <- result$universal_info
  
  cat("Precomputed R²:", universal_info$deconvolution_quality$mean_r_squared, "\n")
}
```

## Format Conversion

### Automatic Format Detection

```{r auto-detect, eval=FALSE}
# Automatically detect and convert format
physeq <- convert_to_phyloseq("my_data.biom")
physeq <- convert_to_phyloseq("feature_table.qza")
physeq <- convert_to_phyloseq("metaphlan_profile.txt")
```

### BIOM Format

Convert BIOM files (from QIIME, mothur, etc.):

```{r biom-convert, eval=FALSE}
# Basic BIOM conversion
physeq <- biom_to_phyloseq("otu_table.biom")

# With sample metadata
physeq <- biom_to_phyloseq(
  biom_file = "otu_table.biom",
  sample_metadata = "sample_data.txt",
  tree_file = "tree.nwk"
)

# From QIIME2 exported BIOM
physeq <- biom_to_phyloseq(
  biom_file = "feature-table.biom",
  taxonomy_file = "taxonomy.tsv",
  tree_file = "tree.nwk",
  metadata_file = "sample-metadata.tsv"
)
```

### QIIME2 Artifacts

Work directly with QIIME2 `.qza` files:

```{r qiime2-convert, eval=FALSE}
# Single artifact
physeq <- qiime2_to_phyloseq("feature_table.qza")

# Multiple artifacts
physeq <- merge_qiime2_artifacts(
  feature_table = "table.qza",
  taxonomy = "taxonomy.qza",
  tree = "rooted-tree.qza",
  metadata = "metadata.tsv"
)

# With filtering
physeq <- qiime2_to_phyloseq(
  qza_file = "table.qza",
  min_reads = 1000,      # Filter samples with < 1000 reads
  min_prevalence = 0.05  # Filter taxa in < 5% of samples
)
```

### MetaPhlAn Profiles

Convert MetaPhlAn taxonomic profiles:

```{r metaphlan-convert, eval=FALSE}
# Single sample
physeq <- metaphlan_to_phyloseq("sample1_profile.txt")

# Multiple samples
files <- list.files("metaphlan_output", pattern = "*_profile.txt", 
                   full.names = TRUE)
physeq <- metaphlan_to_phyloseq(
  profile_files = files,
  metadata_file = "sample_metadata.csv"
)

# With taxonomic level filtering
physeq <- metaphlan_to_phyloseq(
  profile_files = files,
  tax_level = "species",  # Only keep species level
  min_abundance = 0.01    # Filter < 1% abundance
)
```

### Other Formats

```{r other-formats, eval=FALSE}
# Tab-delimited OTU table
physeq <- convert_to_phyloseq(
  "otu_table.tsv",
  format = "tsv",
  taxonomy_file = "taxonomy.tsv",
  sample_data_file = "metadata.tsv"
)

# CSV format
physeq <- convert_to_phyloseq(
  "abundance_table.csv",
  format = "csv",
  taxa_are_rows = TRUE
)

# From data frames
otu_df <- read.csv("otus.csv", row.names = 1)
tax_df <- read.csv("taxonomy.csv", row.names = 1)
physeq <- convert_to_phyloseq(
  otu_table = otu_df,
  taxonomy = tax_df,
  format = "data.frame"
)
```

## Interactive Dataset Browser

### Using the Shiny Interface

```{r shiny-browser, eval=FALSE}
# Launch the interactive explorer
launch_diversity_explorer()

# The Dataset Browser tab provides:
# - Visual cards for each dataset
# - Filtering by type, source, environment
# - Search functionality
# - One-click loading
# - Detailed metadata display
```

### Programmatic Access

Replicate the browser functionality in code:

```{r programmatic-browser}
# Create a dataset catalog
catalog <- list_available_datasets()

# Add tags information
for (i in 1:nrow(catalog)) {
  info <- get_dataset_info(catalog$id[i])
  catalog$tags[i] <- paste(info$tags, collapse = ", ")
  catalog$has_tree[i] <- info$has_tree
}

# Create interactive table
datatable(
  catalog[, c("name", "type", "samples", "taxa", "has_tree", "source")],
  options = list(pageLength = 5),
  caption = "Available Datasets in diversityGPT"
)
```

## Working with Your Own Data

### Best Practices

1. **Prepare Your Data**:
   - OTU/ASV table (samples as columns, taxa as rows)
   - Taxonomy table (optional but recommended)
   - Sample metadata (optional but recommended)
   - Phylogenetic tree (optional, enables P component)

2. **File Formats**:
   - Use standard formats (BIOM, QIIME2, TSV)
   - Include headers in tabular files
   - Use consistent sample IDs across files

3. **Quality Control**:
   ```{r qc-example, eval=FALSE}
   # Check data after loading
   physeq <- convert_to_phyloseq("my_data.biom")
   
   # Basic QC
   cat("Samples:", nsamples(physeq), "\n")
   cat("Taxa:", ntaxa(physeq), "\n")
   cat("Sparsity:", sum(otu_table(physeq) == 0) / length(otu_table(physeq)), "\n")
   
   # Filter low-quality samples/taxa
   physeq_filtered <- phyloseq::filter_taxa(physeq, function(x) sum(x > 0) > 5, TRUE)
   physeq_filtered <- phyloseq::prune_samples(sample_sums(physeq_filtered) > 1000, 
                                              physeq_filtered)
   ```

### Creating a Dataset Entry

Add your dataset to the registry:

```{r create-entry, eval=FALSE}
# Create a dataset entry
my_dataset <- list(
  id = "my_study",
  name = "My Microbiome Study",
  type = "16S",
  source = "custom",
  samples = nsamples(physeq),
  taxa = ntaxa(physeq),
  has_tree = !is.null(phy_tree(physeq, errorIfNULL = FALSE)),
  has_metadata = !is.null(sample_data(physeq, errorIfNULL = FALSE)),
  description = "Time series of gut microbiome during treatment",
  citation = "Your Name et al. (2025) Journal Name",
  tags = c("gut", "time series", "treatment", "16S V4"),
  file_path = "path/to/your/data.rds"
)

# Save phyloseq object
saveRDS(physeq, "path/to/your/data.rds")
```

## Batch Processing

### Processing Multiple Datasets

```{r batch-process, eval=FALSE}
# Get all gut datasets
gut_datasets <- search_datasets("gut")

# Process each dataset
results <- list()
for (i in 1:nrow(gut_datasets)) {
  cat("Processing:", gut_datasets$name[i], "\n")
  
  # Load dataset
  physeq <- load_dataset(gut_datasets$id[i])
  
  # Extract universal information
  universal_info <- extract_universal_information(physeq)
  
  # Store results
  results[[gut_datasets$id[i]]] <- list(
    dataset = gut_datasets$name[i],
    quality = universal_info$deconvolution_quality$mean_r_squared,
    n_samples = nsamples(physeq),
    components = universal_info$deconvolution_quality$components_present
  )
}

# Summarize results
summary_df <- do.call(rbind, lapply(results, as.data.frame))
print(summary_df)
```

### Parallel Processing

```{r parallel, eval=FALSE}
# Use parallel processing for large datasets
library(parallel)

# Set up cluster
cl <- makeCluster(detectCores() - 1)
clusterEvalQ(cl, {
  library(diversityGPT)
  library(phyloseq)
})

# Process datasets in parallel
dataset_ids <- c("dataset1", "dataset2", "dataset3")
results <- parLapply(cl, dataset_ids, function(id) {
  physeq <- load_dataset(id)
  extract_universal_information(physeq)
})

stopCluster(cl)
```

## Format Conversion Tips

### Handling Large Files

```{r large-files, eval=FALSE}
# For large BIOM files, use streaming
physeq <- biom_to_phyloseq(
  biom_file = "large_table.biom",
  chunk_size = 10000  # Process in chunks
)

# For multiple QIIME2 artifacts, process sequentially
artifacts <- list.files("qiime2_output", pattern = "*.qza")
for (artifact in artifacts) {
  # Process and save each
  physeq <- qiime2_to_phyloseq(artifact)
  saveRDS(physeq, sub(".qza", ".rds", artifact))
}
```

### Troubleshooting Common Issues

1. **Memory Issues**:
   ```{r memory, eval=FALSE}
   # Reduce data size before processing
   physeq_small <- prune_taxa(taxa_sums(physeq) > 10, physeq)
   ```

2. **Format Detection Fails**:
   ```{r format-fix, eval=FALSE}
   # Explicitly specify format
   physeq <- convert_to_phyloseq("ambiguous_file.txt", format = "tsv")
   ```

3. **Missing Metadata**:
   ```{r metadata-fix, eval=FALSE}
   # Add metadata after conversion
   physeq <- biom_to_phyloseq("table.biom")
   metadata <- read.csv("metadata.csv", row.names = 1)
   sample_data(physeq) <- metadata
   ```

## Next Steps

- See `vignette("caching-guide")` for performance optimization
- Check the [Shiny app documentation](https://diversitygpt.github.io/shiny) for interactive analysis
- Explore precomputed datasets for instant analysis

The dataset management system in diversityGPT makes it easy to work with microbiome data from any source, ensuring you can focus on analysis rather than data wrangling.