---
title: "Ecological Intelligence: From Patterns to Understanding"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ecological Intelligence: From Patterns to Understanding}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

The diversityGPT package revolutionizes microbiome analysis by combining rigorous mathematical frameworks with ecological intelligence. This vignette demonstrates how to transform diversity patterns into ecological understanding using our comprehensive suite of tools.

## The Ecological Intelligence Pipeline

Our ecological intelligence features follow a logical workflow:

1. **Pattern Detection**: Extract universal information components (R, E, P, S)
2. **Mechanism Identification**: Detect assembly processes driving community structure
3. **Hypothesis Generation**: Create testable predictions about ecological processes
4. **Literature Integration**: Connect findings to existing knowledge
5. **AI-Powered Synthesis**: Generate comprehensive ecological insights

# Quick Start: Complete Workflow

```{r quick-start, eval=FALSE}
library(diversityGPT)

# Load your microbiome data
data("GlobalPatterns") # Example dataset
physeq <- GlobalPatterns

# Run complete ecological intelligence analysis
eco_intelligence <- run_ecological_intelligence(
  physeq,
  environmental_data = sample_data(physeq),
  study_context = list(
    environment = "soil_microbiome",
    research_question = "What drives microbial diversity in different soil types?"
  )
)

# View comprehensive results
print(eco_intelligence)
```

# Step-by-Step Analysis

## Step 1: Extract Universal Information Components

First, we decompose diversity into its fundamental components:

```{r extract-components}
library(diversityGPT)

# Create example data with environmental gradients
set.seed(123)
demo_data <- create_demo_phyloseq(n_samples = 30, n_taxa = 50)

# Add environmental data
env_data <- data.frame(
  pH = rnorm(30, mean = 6.5, sd = 1),
  temperature = rnorm(30, mean = 25, sd = 5),
  moisture = runif(30, min = 10, max = 90)
)
sample_data(demo_data) <- sample_data(env_data)

# Extract universal components
universal_info <- extract_universal_information(demo_data)
print(universal_info)

# Visualize components
plot(universal_info, type = "components")
```

## Step 2: Detect Assembly Mechanisms

Identify the ecological processes shaping your community:

```{r detect-mechanisms}
# Detect assembly mechanisms
mechanisms <- detect_assembly_mechanisms(
  universal_info,
  environmental_data = env_data,
  detect_methods = c("environmental_filtering", "competitive_exclusion", 
                    "neutral_drift", "dispersal_limitation")
)

print(mechanisms)

# Which mechanisms are significant?
significant_mechanisms <- get_significant_mechanisms(mechanisms)
```

## Step 3: Generate Testable Hypotheses

Transform patterns into scientific hypotheses:

```{r generate-hypotheses}
# Generate ecological hypotheses
hypotheses <- generate_ecological_hypotheses(
  patterns = list(diversity = universal_info, assembly = mechanisms),
  study_context = list(
    environment = "soil_microbiome",
    organism = "bacteria",
    temporal_scale = "seasonal",
    spatial_scale = "landscape"
  ),
  hypothesis_types = c("mechanistic", "predictive", "experimental"),
  n_hypotheses = 5
)

# View top hypotheses
print_top_hypotheses(hypotheses, n = 3)
```

## Step 4: Search and Integrate Literature

Connect your findings to existing knowledge:

```{r literature-search}
# Search relevant literature
literature <- search_literature(
  query = "soil microbiome pH temperature assembly",
  databases = c("pubmed", "crossref"),
  max_results = 20,
  filter_criteria = list(
    year_min = 2020,
    relevance_threshold = 0.6
  )
)

# View relevant papers
print(literature)

# Identify research gaps
gaps <- identify_research_gaps(literature, hypotheses)
```

## Step 5: AI-Powered Synthesis (Optional)

If you have configured an LLM API, get comprehensive ecological insights:

```{r llm-analysis, eval=FALSE}
# Configure LLM (only needed once)
configure_llm(
  provider = "anthropic",  # or "openai"
  api_key = "your-api-key"
)

# Run multi-step analysis
ai_insights <- llm_multi_step_analysis(
  data = list(
    diversity = universal_info,
    mechanisms = mechanisms,
    hypotheses = hypotheses,
    literature = literature
  ),
  context = list(
    study_system = "soil microbiome",
    research_question = "How do pH and temperature interact to structure soil bacterial communities?",
    expertise_level = "expert"
  ),
  reasoning_depth = "deep"
)

# View AI insights
cat(ai_insights$synthesis)
```

# Real-World Case Studies

## Case Study 1: Environmental Filtering in Extreme Environments

```{r case-study-1}
# Simulate extreme environment data
extreme_data <- create_demo_phyloseq(n_samples = 20, n_taxa = 30)

# Add extreme environmental gradients
extreme_env <- data.frame(
  pH = c(rep(3, 10), rep(9, 10)),  # Extreme pH
  temperature = c(rep(5, 10), rep(45, 10)),  # Extreme temperatures
  site = factor(c(rep("acidic_cold", 10), rep("alkaline_hot", 10)))
)
sample_data(extreme_data) <- sample_data(extreme_env)

# Analyze assembly mechanisms
extreme_universal <- extract_universal_information(extreme_data)
extreme_mechanisms <- detect_assembly_mechanisms(
  extreme_universal,
  environmental_data = extreme_env[, c("pH", "temperature")]
)

# Results show strong environmental filtering
print(extreme_mechanisms)
```

## Case Study 2: Competitive Exclusion in Nutrient-Rich Environments

```{r case-study-2}
# Simulate competitive environment
competitive_data <- create_demo_phyloseq(n_samples = 25, n_taxa = 40)

# High richness, low evenness pattern (competitive exclusion signature)
otu_table(competitive_data)[1:5, ] <- otu_table(competitive_data)[1:5, ] * 10

# Analyze patterns
competitive_universal <- extract_universal_information(competitive_data)
competitive_mechanisms <- detect_assembly_mechanisms(competitive_universal)

# Generate hypotheses about competition
competitive_hypotheses <- generate_ecological_hypotheses(
  patterns = list(diversity = competitive_universal),
  study_context = list(
    environment = "nutrient_rich",
    expected_process = "competition"
  )
)

print(competitive_hypotheses)
```

# Advanced Features

## Custom Assembly Mechanism Detection

You can define custom assembly mechanisms:

```{r custom-mechanisms, eval=FALSE}
# Define custom mechanism detector
custom_detector <- function(universal_info, env_data) {
  # Your custom logic here
  # Return detection result with confidence
}

# Add to detection
mechanisms <- detect_assembly_mechanisms(
  universal_info,
  custom_detectors = list(
    my_mechanism = custom_detector
  )
)
```

## Hypothesis Testing Framework

Test generated hypotheses with your data:

```{r hypothesis-testing, eval=FALSE}
# Select hypothesis to test
test_hypothesis <- hypotheses$hypotheses[[1]]

# Design experiment
experiment <- design_experiment(
  hypothesis = test_hypothesis,
  current_data = demo_data,
  feasibility_constraints = list(
    max_samples = 50,
    time_points = 5
  )
)

# Evaluate hypothesis with new data
results <- test_ecological_hypothesis(
  hypothesis = test_hypothesis,
  test_data = new_data,
  statistical_tests = c("permanova", "correlation")
)
```

# Visualization Gallery

## Assembly Mechanism Summary

```{r viz-mechanisms, fig.height=6}
# Create mechanism summary plot
plot_mechanism_summary <- function(mechanisms) {
  # Extract confidence scores
  mech_data <- data.frame(
    mechanism = names(mechanisms$mechanisms),
    confidence = sapply(mechanisms$mechanisms, function(x) x$confidence),
    detected = sapply(mechanisms$mechanisms, function(x) x$detected)
  )
  
  # Create bar plot
  library(ggplot2)
  ggplot(mech_data, aes(x = mechanism, y = confidence, fill = detected)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = c("FALSE" = "gray", "TRUE" = "darkgreen")) +
    coord_flip() +
    theme_minimal() +
    labs(title = "Assembly Mechanism Detection",
         x = "Mechanism",
         y = "Confidence Score")
}

# Plot if mechanisms exist
if (exists("mechanisms")) {
  plot_mechanism_summary(mechanisms)
}
```

## Hypothesis Network

```{r viz-hypotheses, fig.height=7}
# Visualize hypothesis relationships
if (exists("hypotheses") && length(hypotheses$hypotheses) > 0) {
  plot(hypotheses, type = "network")
}
```

# Best Practices

## 1. Start with Quality Data
- Ensure adequate sampling depth
- Include relevant environmental metadata
- Use appropriate controls

## 2. Validate Mechanisms
- Use multiple detection methods
- Check consistency across subsets
- Compare with literature expectations

## 3. Prioritize Hypotheses
- Focus on testable predictions
- Consider feasibility constraints
- Build on existing knowledge

## 4. Iterate and Refine
- Test initial hypotheses
- Update understanding
- Generate new predictions

# Troubleshooting

## Common Issues

**No mechanisms detected:**
```{r troubleshoot-1, eval=FALSE}
# Check data properties
check_data_properties(universal_info)

# Lower confidence threshold
mechanisms <- detect_assembly_mechanisms(
  universal_info,
  confidence_threshold = 0.5  # Default is 0.7
)
```

**Poor hypothesis quality:**
```{r troubleshoot-2, eval=FALSE}
# Provide more context
hypotheses <- generate_ecological_hypotheses(
  patterns = patterns,
  study_context = list(
    environment = "specific_ecosystem",
    organism = "target_taxa",
    known_drivers = c("temperature", "pH"),
    research_goals = "understand seasonal variation"
  )
)
```

# Conclusion

The ecological intelligence features in diversityGPT provide a complete pipeline from pattern detection to ecological understanding. By combining rigorous mathematical frameworks with ecological theory and AI assistance, researchers can:

1. Identify assembly mechanisms with statistical confidence
2. Generate testable hypotheses automatically
3. Connect findings to existing literature
4. Synthesize comprehensive ecological insights

This transforms microbiome analysis from descriptive statistics to mechanistic understanding, accelerating scientific discovery.

# Further Reading

- Assembly Mechanisms: See `?detect_assembly_mechanisms` for details
- Hypothesis Generation: See `?generate_ecological_hypotheses` for options
- Literature Integration: See `?search_literature` for database options
- Universal Information: See `vignette("universal_information_theory")`

# Session Info

```{r session}
sessionInfo()
```