---
title: "Getting Started with diversityGPT"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with diversityGPT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

Welcome to **diversityGPT** - the revolutionary R package that transforms how we analyze microbiome diversity. This vignette will guide you through the core concepts and basic usage of the package.

### What Makes diversityGPT Different?

Traditional diversity analysis treats each metric (Shannon, Simpson, Chao1, etc.) as an independent measurement. When different metrics disagree, researchers face "decision confusion" - which metric should they trust?

**diversityGPT solves this by revealing that ALL diversity metrics are mathematically related through information theory.** Every metric can be decomposed into fundamental components:

- **R (Richness)**: Information about species count
- **E (Evenness)**: Information about distribution uniformity  
- **P (Phylogenetic)**: Information about evolutionary relationships
- **S (Spatial)**: Information about geographic patterns

This universal framework enables:
1. Converting between ANY diversity metrics
2. Predicting missing metrics with high accuracy (R² > 0.9)
3. Resolving conflicts when metrics disagree
4. Standardizing metrics across different studies

## Installation and Setup

```{r setup, eval=FALSE}
# Install from GitHub
devtools::install_github("shandley/diversityGPT")

# Load the package
library(diversityGPT)
library(phyloseq)
```

```{r load-quiet, echo=FALSE}
library(diversityGPT)
library(phyloseq)
```

### Optional: AI Integration

For AI-powered ecological interpretation, add your API key to `.Renviron`:

```bash
ANTHROPIC_API_KEY=your-key-here
# or
OPENAI_API_KEY=your-key-here
```

## Basic Workflow

Let's walk through a complete analysis using the GlobalPatterns dataset:

### 1. Load Your Data

```{r load-data}
# Option 1: Use built-in phyloseq data
data(GlobalPatterns)

# Option 2: Load from our dataset collection
# physeq <- load_dataset("enterotype")

# Option 3: Convert from common formats
# physeq <- convert_to_phyloseq("my_data.biom")

# Quick look at the data
print(GlobalPatterns)
```

### 2. Extract Universal Information Components

This is the core innovation of diversityGPT - decomposing all diversity information into universal components:

```{r extract-info}
# Extract universal information
universal_info <- extract_universal_information(
  GlobalPatterns,
  groups = "SampleType"  # Optional grouping variable
)

# View the results
print(universal_info)
```

The output shows:
- **Overall Quality**: How well the universal framework captures your data's diversity
- **Mean R²**: Average accuracy of metric predictions
- **Components Present**: Which information types are in your data

### 3. Explore the Information Components

Let's visualize how diversity information is distributed:

```{r plot-components, fig.cap="Distribution of diversity information across R, E, P, S components"}
# Plot information components
plot_information_components(universal_info)
```

Each bar shows how much each sample's diversity comes from:
- **Red (R)**: Species richness
- **Blue (E)**: Evenness of distribution
- **Green (P)**: Phylogenetic diversity (if tree present)
- **Orange (S)**: Spatial patterns (if coordinates present)

### 4. Transform Between Metrics

Now for the magic - convert between ANY diversity metrics:

```{r transform-metrics}
# Say we have Shannon diversity but need Simpson and Chao1
source_metrics <- c(shannon = 2.5)

# Transform to other metrics
predictions <- universal_diversity_transform(
  source_metrics = source_metrics,
  target_metrics = c("simpson", "chao1", "observed"),
  transformation_matrix = universal_info$transformation_matrix
)

# View predictions
print(predictions)
```

### 5. Visualize Metric Relationships

See how all metrics relate to each other:

```{r plot-network, fig.cap="Network showing relationships between diversity metrics"}
# Create interactive network plot
plot_diversity_network(universal_info, interactive = FALSE)
```

Thicker lines indicate stronger relationships between metrics.

### 6. Predict Missing Metrics

What if you need Faith's PD but don't have a phylogenetic tree? Or want Bray-Curtis but only have alpha diversity?

```{r predict-missing}
# Predict missing metrics from available ones
available_metrics <- c(shannon = 2.3, observed = 150)

missing_predictions <- predict_missing_diversity_metrics(
  available_metrics = available_metrics,
  phyloseq_reference = GlobalPatterns  # Uses this as reference
)

head(missing_predictions)
```

## Interactive Analysis

For a more comprehensive analysis experience, launch the Shiny app:

```{r shiny, eval=FALSE}
# Launch the interactive explorer
launch_diversity_explorer()

# Or try the component explorer
launch_component_explorer()
```

The interactive explorer provides:
- Real-time metric transformation
- Dataset browsing and loading
- Interactive visualizations
- AI-powered interpretation
- Export capabilities

## Common Use Cases

### Use Case 1: Resolving Metric Conflicts

When Shannon shows significant differences but Simpson doesn't:

```{r conflict-resolution}
# First calculate diversity metrics
div_results <- calculate_diversity(GlobalPatterns)

# Then calculate consensus across metrics
consensus_result <- consensus_diversity(
  div_results,
  method = "information_weighted"
)

# The consensus considers all metrics weighted by their information content
print(head(consensus_result$consensus_value))
```

### Use Case 2: Cross-Study Comparison

Different studies used different metrics? Standardize them:

```{r cross-study, eval=FALSE}
# Study 1 used Shannon and Simpson
study1_universal <- extract_universal_information(study1_data)

# Study 2 used Chao1 and Faith's PD  
study2_universal <- extract_universal_information(study2_data)

# Now both are comparable through R, E, P, S components!
```

### Use Case 3: Ecological Interpretation

Get AI-powered insights about your patterns:

```{r ai-interpretation, eval=FALSE}
# Get ecological interpretation
interpretation <- interpret_diversity(
  universal_info,
  context = list(
    environment = "human_gut",
    condition = "antibiotic_treatment",
    hypothesis = "We expect reduced diversity after antibiotics"
  )
)

print(interpretation)
```

## Key Concepts Summary

1. **Universal Information Framework**: All diversity metrics share common information components (R, E, P, S)

2. **Transformation Matrix**: Mathematical relationships that enable metric conversion

3. **Quality Metrics**: R² values tell you how reliable each transformation is

4. **Consensus Methods**: Resolve conflicts by considering all metrics together

5. **AI Integration**: Get ecological interpretation of mathematical patterns

## Next Steps

- Explore the `vignette("universal-transformation")` for mathematical details
- See `vignette("dataset-management")` for data handling options  
- Check `vignette("caching-guide")` for performance optimization
- Try the interactive explorer with your own data!

## Getting Help

- **Documentation**: Type `?function_name` for any function
- **GitHub Issues**: https://github.com/shandley/diversityGPT/issues
- **Discussions**: https://github.com/shandley/diversityGPT/discussions

Remember: diversityGPT doesn't replace your ecological knowledge - it enhances it by revealing hidden mathematical relationships in your data!