---
title: "Introduction to diversityGPT"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to diversityGPT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = TRUE
)
```

```{r setup}
library(diversityGPT)
library(phyloseq)
```

## Overview

diversityGPT revolutionizes microbiome diversity analysis by solving the "metric confusion problem" - when different diversity metrics give conflicting results. The package provides:

1. **Universal Metric Transformation**: Convert between ANY diversity metrics
2. **AI-Powered Interpretation**: Understand ecological meaning of patterns
3. **Consensus Analysis**: Resolve conflicts between metrics
4. **Interactive Exploration**: Visualize and explore diversity patterns

## Quick Start

### Basic Analysis

```{r basic-analysis}
# Load example microbiome data
data(GlobalPatterns)

# Run comprehensive diversity analysis
results <- diversity_suite(
  phyloseq_obj = GlobalPatterns,
  metrics = c("shannon", "simpson", "chao1", "faith_pd"),
  consensus_method = "adaptive"
)

# View results
print(results)
```

### Visualize Diversity Patterns

```{r visualize}
# Plot diversity metrics
plot(results, type = "metrics")

# Show consensus vs individual metrics
plot(results, type = "consensus")
```

## Universal Information Framework

The core innovation of diversityGPT is the ability to decompose any diversity metric into fundamental components:

```{r universal-info}
# Extract universal information
universal_info <- extract_universal_information(GlobalPatterns)

# View information components
head(universal_info$information_components)

# Visualize component breakdown
plot(universal_info, type = "components")
```

## Metric Transformation

Transform between any diversity metrics:

```{r transformation}
# Example: You have Shannon, want Simpson and Chao1
source_metrics <- data.frame(
  shannon = c(2.1, 2.5, 1.8, 2.3),
  row.names = paste0("Sample", 1:4)
)

# Transform to other metrics
transformed <- universal_diversity_transform(
  source_metrics = source_metrics,
  target_metrics = c("simpson", "chao1"),
  transformation_matrix = universal_info$transformation_matrix
)

print(transformed)
```

## AI-Powered Interpretation

Get ecological insights from your diversity patterns:

```{r ai-interpretation, eval=FALSE}
# Configure LLM (requires API key)
configure_llm(
  provider = "anthropic",
  api_key = "your-api-key"
)

# Get AI interpretation
interpretation <- interpret_diversity(
  results,
  context = list(
    environment = "human_gut",
    condition = "healthy_vs_disease"
  )
)

print(interpretation$summary)
print(interpretation$hypotheses)
```

## Consensus Analysis

When metrics disagree, use consensus algorithms:

```{r consensus}
# Run consensus analysis
consensus <- consensus_diversity(
  phyloseq_obj = GlobalPatterns,
  metrics = c("shannon", "simpson", "observed"),
  method = "information_theoretic"
)

# Plot confidence intervals
plot(consensus, type = "bootstrap")
```

## Interactive Exploration

Launch the interactive Shiny application:

```{r shiny, eval=FALSE}
# Launch interactive explorer
launch_diversity_explorer()
```

The Shiny app provides:
- Real-time metric transformation
- Interactive visualizations
- AI-powered insights
- Data upload and export

## Working with Your Data

### From phyloseq

```{r your-data-phyloseq, eval=FALSE}
# Your phyloseq object
my_data <- phyloseq(
  otu_table(otu_matrix, taxa_are_rows = TRUE),
  sample_data(metadata),
  tax_table(taxonomy)
)

# Analyze
results <- diversity_suite(my_data)
```

### From BIOM Format

```{r your-data-biom, eval=FALSE}
# Convert BIOM to phyloseq
my_data <- biom_to_phyloseq("path/to/file.biom")

# Analyze
results <- diversity_suite(my_data)
```

### From CSV Files

```{r your-data-csv, eval=FALSE}
# Load OTU table and metadata
otu_table <- read.csv("otu_table.csv", row.names = 1)
metadata <- read.csv("metadata.csv", row.names = 1)

# Create phyloseq object
my_data <- phyloseq(
  otu_table(as.matrix(otu_table), taxa_are_rows = TRUE),
  sample_data(metadata)
)

# Analyze
results <- diversity_suite(my_data)
```

## API for Programmatic Access

For advanced users and pipeline integration:

```{r api-example}
# Use the unified API
result <- diversityGPT_api(
  "analyze",
  phyloseq_obj = GlobalPatterns,
  components = c("universal", "mechanisms")
)

# Check status
if (is_api_success(result)) {
  data <- get_api_data(result)
  print(names(data))
}
```

## Next Steps

1. Read the `vignette("universal-framework")` to understand the mathematical foundations
2. Explore `vignette("ecological-intelligence")` for AI-powered features
3. Check `vignette("advanced-analysis")` for complex workflows
4. Join our community at https://github.com/shandley/diversityGPT

## Getting Help

```{r help, eval=FALSE}
# Package documentation
?diversityGPT

# Function help
?extract_universal_information
?universal_diversity_transform

# Browse all vignettes
browseVignettes("diversityGPT")
```

## Citation

If you use diversityGPT in your research, please cite:

```{r citation}
citation("diversityGPT")
```