---
title: "Robust Taxa Indicator Analysis: The Most Rigorous Approach"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Robust Taxa Indicator Analysis: The Most Rigorous Approach}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  warning = FALSE,
  message = FALSE
)
```

# Introduction

The `robust_taxa_indicators()` function represents the pinnacle of taxa indicator analysis, integrating four complementary mathematical approaches to provide the most statistically rigorous identification of indicator taxa available. This vignette demonstrates how to use this revolutionary framework that combines:

1. **Null Model Validation**: Test against random expectations
2. **Information Theory**: Quantify taxa-component relationships  
3. **Game Theory (Shapley Values)**: Fair attribution of contributions
4. **Bootstrap Uncertainty**: Confidence intervals and stability assessment

# Why Robust Taxa Indicators?

Traditional approaches to identifying "driver" taxa often rely on simple correlation or abundance metrics, which can be misleading. Our robust approach addresses critical issues:

- **Multiple Testing**: Validates findings against null models
- **Fair Attribution**: Uses game theory for unbiased contribution assessment
- **Information Content**: Quantifies actual information transfer
- **Uncertainty Quantification**: Provides confidence intervals
- **Method Agreement**: Consensus across different mathematical frameworks

# Quick Start

```{r quickstart}
library(diversityGPT)

# Load example data
data("GlobalPatterns", package = "phyloseq")
physeq <- GlobalPatterns

# Subset for demonstration (faster computation)
physeq_subset <- prune_taxa(taxa_sums(physeq) > 1000, physeq)
physeq_subset <- prune_samples(sample_sums(physeq_subset) > 5000, physeq_subset)

# Run comprehensive robust analysis
robust_results <- robust_taxa_indicators(
  physeq_subset,
  methods = c("null_models", "mutual_info", "shapley", "bootstrap"),
  top_n = 10,
  n_permutations = 99,   # Increase for publication (999+)
  n_bootstrap = 99,      # Increase for publication (999+)
  confidence_level = 0.95,
  verbose = TRUE
)

# View results
print(robust_results)
```

# Step-by-Step Analysis

## Step 1: Data Preparation

First, ensure your data is properly formatted and filtered:

```{r data-prep}
# Create a well-structured dataset
set.seed(123)
demo_physeq <- create_demo_phyloseq(n_samples = 30, n_taxa = 50)

# Add meaningful sample metadata
sample_data(demo_physeq) <- sample_data(
  data.frame(
    condition = rep(c("control", "treatment"), each = 15),
    timepoint = rep(c("T0", "T1", "T2"), 10),
    row.names = sample_names(demo_physeq)
  )
)

# Filter low-abundance taxa (optional but recommended)
demo_physeq <- prune_taxa(taxa_sums(demo_physeq) > 10, demo_physeq)
```

## Step 2: Extract Universal Information Components

Before identifying indicators, extract the mathematical components:

```{r extract-components}
# Extract R, E, P, S components
universal_info <- extract_universal_information(demo_physeq)

# These components represent:
# R = Richness information
# E = Evenness information  
# P = Phylogenetic information
# S = Spatial/sample distribution information

print(universal_info$components_summary)
```

## Step 3: Run Robust Analysis

Now perform the comprehensive indicator analysis:

```{r robust-analysis}
# Full robust analysis with all methods
robust_indicators <- robust_taxa_indicators(
  demo_physeq,
  components = universal_info,  # Use pre-extracted components
  methods = c("null_models", "mutual_info", "shapley", "bootstrap"),
  top_n = 15,
  n_permutations = 199,
  n_bootstrap = 199,
  confidence_level = 0.95,
  parallel = FALSE,  # Set TRUE for large datasets
  seed = 42,
  verbose = TRUE
)
```

## Step 4: Examine Results

### Consensus Indicators

The consensus rankings combine evidence from all methods:

```{r consensus-results}
# View top richness indicators
richness_indicators <- robust_indicators$consensus_indicators$richness
print(head(richness_indicators, 10))

# Interpretation of columns:
# - consensus_score: Combined evidence across methods (higher = stronger indicator)
# - n_methods: Number of methods identifying this taxon
# - reliability: Confidence in this taxon as an indicator (0-1)
# - final_uncertainty: Uncertainty estimate from bootstrap
```

### Method Agreement

Check how well different methods agree:

```{r method-agreement}
# Overall agreement between methods
cat("Overall method agreement:", 
    round(robust_indicators$method_agreement$overall$mean_agreement * 100, 1), "%\n")

# Component-specific agreement
for (comp in c("richness", "evenness", "phylogenetic", "spatial")) {
  if (!is.null(robust_indicators$method_agreement[[comp]])) {
    cat(comp, "agreement:", 
        round(robust_indicators$method_agreement[[comp]]$mean_agreement * 100, 1), "%\n")
  }
}
```

### Reliability Assessment

Understand the confidence in your results:

```{r reliability}
# Overall reliability score
cat("Overall reliability score:", 
    round(robust_indicators$reliability_scores$overall_score, 3), "\n")
cat("Interpretation:", 
    robust_indicators$summary_report$reliability$interpretation, "\n")

# Method-specific reliability
if (!is.null(robust_indicators$reliability_scores$null_models)) {
  cat("\nNull model significance rates:\n")
  print(robust_indicators$reliability_scores$null_models$significance_rate)
}
```

# Visualization

## Consensus Indicator Plot

```{r plot-consensus, fig.height=8}
# Plot top consensus indicators for richness
plot(robust_indicators, 
     type = "consensus", 
     component = "richness", 
     top_n = 10)
```

## Method Comparison

```{r plot-comparison, fig.height=6}
# Compare rankings across methods
plot(robust_indicators, 
     type = "comparison", 
     top_n = 8)
```

## Reliability Visualization

```{r plot-reliability, fig.height=6}
# Visualize reliability metrics
plot(robust_indicators, 
     type = "reliability")
```

# Understanding Each Method

## 1. Null Model Validation

Tests whether observed indicator values exceed random expectation:

```{r null-model-details}
# Extract null model results
null_results <- robust_indicators$null_model_results

# Check p-values for richness indicators
if (!is.null(null_results)) {
  richness_pvals <- null_results$p_values$richness
  significant_taxa <- names(richness_pvals)[richness_pvals < 0.05]
  
  cat("Significant richness indicators (p < 0.05):", 
      length(significant_taxa), "out of", length(richness_pvals), "\n")
  
  # Effect sizes for significant taxa
  if (length(significant_taxa) > 0) {
    effect_sizes <- null_results$effect_sizes$richness[significant_taxa]
    cat("Top 5 effect sizes:\n")
    print(sort(effect_sizes, decreasing = TRUE)[1:5])
  }
}
```

## 2. Mutual Information Analysis

Quantifies information shared between taxa and diversity components:

```{r mutual-info-details}
# Extract mutual information results
mi_results <- robust_indicators$mutual_info_results

if (!is.null(mi_results)) {
  # Total information content
  cat("Total information by component:\n")
  print(mi_results$summary_stats$total_information)
  
  # Top information-rich taxa for richness
  top_mi_taxa <- head(mi_results$taxa_rankings$richness, 5)
  print(top_mi_taxa)
}
```

## 3. Shapley Value Attribution

Fair game-theoretic attribution of contributions:

```{r shapley-details}
# Extract Shapley results
shapley_results <- robust_indicators$shapley_results

if (!is.null(shapley_results)) {
  # Check fairness properties
  cat("Shapley fairness axioms:\n")
  cat("- Efficiency satisfied:", 
      all(sapply(shapley_results$fairness_properties, 
                function(x) x$efficiency$satisfied)), "\n")
  
  # Top Shapley contributors
  shapley_rich <- shapley_results$shapley_matrix[, "richness"]
  top_shapley <- sort(shapley_rich, decreasing = TRUE)[1:5]
  cat("\nTop 5 Shapley contributors to richness:\n")
  print(top_shapley)
}
```

## 4. Bootstrap Uncertainty

Confidence intervals and stability assessment:

```{r bootstrap-details}
# Extract bootstrap results
boot_results <- robust_indicators$bootstrap_results

if (!is.null(boot_results)) {
  # Method stability across bootstrap samples
  cat("Method stability scores:\n")
  print(boot_results$reliability_assessment$method_stability)
  
  # Confidence intervals for top indicators
  if (!is.null(boot_results$consensus_rankings$richness)) {
    boot_consensus <- boot_results$consensus_rankings$richness
    top_5 <- head(boot_consensus, 5)
    
    cat("\nTop 5 indicators with 95% confidence intervals:\n")
    for (i in 1:nrow(top_5)) {
      cat(sprintf("%s: %.3f [%.3f, %.3f]\n",
                 top_5$taxon[i],
                 top_5$mean_rank[i],
                 top_5$ci_lower[i],
                 top_5$ci_upper[i]))
    }
  }
}
```

# Advanced Usage

## Custom Method Selection

Run with specific method combinations:

```{r custom-methods}
# Only null models and mutual information
fast_robust <- robust_taxa_indicators(
  demo_physeq,
  methods = c("null_models", "mutual_info"),
  top_n = 10,
  n_permutations = 99,
  verbose = FALSE
)

# Only Shapley values with bootstrap
shapley_boot <- robust_taxa_indicators(
  demo_physeq,
  methods = c("shapley", "bootstrap"),
  top_n = 8,
  n_bootstrap = 199,
  verbose = FALSE
)
```

## Parallel Processing

For large datasets, enable parallel processing:

```{r parallel-example, eval=FALSE}
# Use multiple cores
library(parallel)
n_cores <- detectCores() - 1

robust_parallel <- robust_taxa_indicators(
  large_physeq,
  methods = c("null_models", "mutual_info", "shapley", "bootstrap"),
  top_n = 20,
  n_permutations = 999,
  n_bootstrap = 999,
  parallel = TRUE,
  n_cores = n_cores,
  verbose = TRUE
)
```

## Group-Specific Analysis

Analyze indicators within specific sample groups:

```{r group-analysis}
# Subset to treatment group only
treatment_samples <- sample_data(demo_physeq)$condition == "treatment"
treatment_physeq <- prune_samples(treatment_samples, demo_physeq)

# Find treatment-specific indicators
treatment_indicators <- robust_taxa_indicators(
  treatment_physeq,
  methods = c("null_models", "mutual_info"),
  top_n = 10,
  n_permutations = 99,
  verbose = FALSE
)

print(treatment_indicators)
```

# Reporting Results

## Generate Comprehensive Report

```{r report-generation, eval=FALSE}
# Create HTML report with all results
report_robust_indicators(
  robust_indicators,
  output_file = "robust_indicators_report.html",
  title = "Comprehensive Taxa Indicator Analysis"
)
```

## Export Results

```{r export-results}
# Export consensus indicators to CSV
write.csv(
  robust_indicators$consensus_indicators$richness,
  "richness_indicators.csv",
  row.names = FALSE
)

# Save complete results object
saveRDS(robust_indicators, "robust_indicators_results.rds")
```

# Best Practices

## 1. Sample Size Considerations
- Minimum 20 samples recommended for reliable results
- More samples improve null model and bootstrap reliability

## 2. Permutation Numbers
- Development: 99-199 permutations for speed
- Publication: 999+ permutations for rigor
- Bootstrap: Match permutation numbers

## 3. Interpreting Results
- **High consensus + low uncertainty** = Most reliable indicators
- **Method disagreement** = Potential method-specific biases
- **Low reliability scores** = Need more data or different approaches

## 4. Computational Resources
- Small datasets (<100 taxa): All methods feasible
- Medium datasets (100-1000 taxa): Consider parallel processing
- Large datasets (>1000 taxa): Use approximation methods for Shapley

# Troubleshooting

## Memory Issues

```{r memory-solutions, eval=FALSE}
# For large datasets, process in chunks
taxa_chunks <- split(taxa_names(large_physeq), 
                    ceiling(seq_along(taxa_names(large_physeq))/100))

chunk_results <- lapply(taxa_chunks, function(chunk) {
  sub_physeq <- prune_taxa(chunk, large_physeq)
  robust_taxa_indicators(sub_physeq, methods = "mutual_info", top_n = 10)
})
```

## Convergence Issues

```{r convergence-check}
# Check if bootstrap has converged
if (!is.null(robust_indicators$bootstrap_results)) {
  convergence <- robust_indicators$bootstrap_results$convergence_assessment
  if (!is.null(convergence)) {
    cat("Bootstrap convergence achieved:", convergence$converged, "\n")
  }
}
```

# Mathematical Details

## Information Theory Foundation

The mutual information between taxon $i$ and component $c$ is:

$$MI(X_i, Y_c) = \sum_{x,y} p(x,y) \log\frac{p(x,y)}{p(x)p(y)}$$

## Shapley Value Calculation

For taxon $i$'s contribution to component $c$:

$$\phi_i = \sum_{S \subseteq N \setminus \{i\}} \frac{|S|!(n-|S|-1)!}{n!}[v(S \cup \{i\}) - v(S)]$$

## Null Model Validation

Effect size calculation:

$$ES_i = \frac{observed_i - mean(null_i)}{sd(null_i)}$$

# Conclusion

The `robust_taxa_indicators()` function provides the most comprehensive and statistically rigorous approach to identifying indicator taxa. By combining multiple mathematical frameworks, it overcomes the limitations of single-method approaches and provides reliable, interpretable results with proper uncertainty quantification.

For questions or issues, please visit: https://github.com/shandley/diversityGPT

# Session Info

```{r session}
sessionInfo()
```