---
title: "Universal Diversity Metric Transformation Framework"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Universal Diversity Metric Transformation Framework}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(diversityGPT)
library(phyloseq)
```

## Introduction

The diversityGPT package introduces a revolutionary mathematical framework that solves a fundamental problem in microbiome analysis: the inability to compare and transform between different diversity metrics. This vignette explains the theoretical foundations and practical applications of the Universal Information Framework.

## The Problem: Diversity Metric Confusion

Microbiome researchers face a critical challenge: different diversity metrics often give conflicting results. For example:

- Shannon diversity might increase while Simpson diversity decreases
- Chao1 (richness) might be high while Faith's PD (phylogenetic diversity) is low
- Different studies use different metrics, making meta-analysis difficult

Traditional approaches treat each metric in isolation, missing the underlying mathematical relationships between them.

## The Solution: Information Theory Decomposition

diversityGPT implements a universal framework based on information theory that decomposes ANY diversity metric into four fundamental components:

1. **R (Richness)**: Information about the number of unique taxa
2. **E (Evenness)**: Information about the distribution of abundances
3. **P (Phylogenetic)**: Information about evolutionary relationships
4. **S (Spatial)**: Information about spatial or functional relationships

## Mathematical Foundation

Every diversity metric can be expressed as:

$$\text{Metric} = w_R \cdot R + w_E \cdot E + w_P \cdot P + w_S \cdot S + \epsilon$$

Where:
- $w_R, w_E, w_P, w_S$ are metric-specific weights
- $\epsilon$ is residual information unique to that metric

This decomposition enables:
- **Transformation**: Convert any metric to any other metric
- **Prediction**: Estimate missing metrics from available ones
- **Standardization**: Compare metrics across different studies

## Basic Usage

### Extract Universal Information

```{r universal-extraction}
# Create demo dataset for fast examples
demo_data <- create_demo_phyloseq(n_samples = 10, n_taxa = 30)

# Extract universal information components
universal_info <- extract_universal_information(
  demo_data,
  include_phylogenetic = FALSE
)

# View the structure
names(universal_info)
```

The `universal_info` object contains:
- **information_components**: R, E, P, S values for each sample
- **transformation_matrix**: Mathematical relationships between metrics
- **deconvolution_quality**: Reliability scores for transformations

### Visualize Information Components

```{r plot-components, fig.height=6}
# Plot the information components
plot(universal_info, type = "components")
```

This visualization shows how each sample's diversity is composed of the four fundamental components.

## Metric Transformation

### Transform Between Metrics

```{r transformation}
# Get sample names from the demo data
sample_names <- sample_names(demo_data)[1:3]

# Example: Transform Shannon to Simpson
source_metrics <- data.frame(
  shannon = c(2.1, 2.5, 1.8),
  row.names = sample_names
)

# Perform transformation
transformed <- universal_diversity_transform(
  source_metrics = source_metrics,
  target_metrics = c("simpson"),
  transformation_matrix = universal_info$transformation_matrix
)

# View results
print(transformed$predicted_metrics)
```

### Assess Transformation Quality

```{r quality-assessment}
# The transformation includes quality metrics
print(transformed$transformation_quality)
```

R² values indicate the reliability of each transformation. Values > 0.8 indicate high reliability.

## Advanced Applications

### Cross-Study Standardization

When comparing studies that used different diversity metrics:

```{r cross-study, eval=FALSE}
# Study 1 used Shannon and Simpson
study1_metrics <- data.frame(
  shannon = c(2.1, 2.3, 2.5),
  simpson = c(0.8, 0.85, 0.9)
)

# Study 2 used Chao1 and Faith's PD
study2_metrics <- data.frame(
  chao1 = c(45, 52, 48),
  faith_pd = c(8.2, 9.1, 8.7)
)

# Standardize to common metrics
study1_standardized <- universal_diversity_transform(
  source_metrics = study1_metrics,
  target_metrics = c("observed", "evenness"),
  transformation_matrix = universal_info$transformation_matrix
)

study2_standardized <- universal_diversity_transform(
  source_metrics = study2_metrics,
  target_metrics = c("observed", "evenness"),
  transformation_matrix = universal_info$transformation_matrix
)

# Now both studies can be compared using the same metrics
```

### Missing Metric Imputation

When some diversity metrics are unavailable:

```{r imputation}
# Use actual sample names from demo data
sample_names <- sample_names(demo_data)[1:3]

# Have Shannon, want to predict other metrics
available_metrics <- data.frame(
  shannon = c(2.1, 2.5, 1.8),
  observed = c(35, 42, 28),
  row.names = sample_names
)

# Predict missing metrics
predicted <- predict_missing_diversity_metrics(
  available_metrics = available_metrics,
  phyloseq_reference = demo_data
)

print(predicted$predicted_metrics)
```

### Relationship Discovery

Explore mathematical relationships between metrics:

```{r relationships}
# Discover relationships
relationships <- discover_metric_relationships(demo_data)

# Print relationship summary
print(relationships$summary)
```

## Theoretical Insights

### Why This Works

The universal framework works because:

1. **Conservation of Information**: Total diversity information is conserved across transformations
2. **Linear Decomposition**: Most diversity metrics are linear combinations of fundamental components
3. **Orthogonal Components**: R, E, P, S capture independent aspects of diversity

### Component Interpretation

- **High R, Low E**: Many rare species (diverse but uneven)
- **Low R, High E**: Few species but evenly distributed
- **High P**: Phylogenetically diverse community
- **High S**: Functionally or spatially structured community

### Limitations

- **Nonlinear Metrics**: Some metrics (e.g., Berger-Parker) have nonlinear relationships
- **Data Requirements**: Transformation quality depends on reference dataset diversity
- **Phylogenetic Info**: P component requires phylogenetic tree

## Best Practices

1. **Use Representative Reference Data**: The transformation matrix quality depends on the diversity of the reference dataset

2. **Check Transformation Quality**: Always examine R² values before trusting transformations

3. **Validate with Known Data**: When possible, validate transformations against datasets where multiple metrics are available

4. **Consider Context**: Some transformations are more reliable within similar ecosystem types

## Conclusion

The Universal Information Framework in diversityGPT provides:
- Mathematical foundation for diversity metric relationships
- Practical tools for metric transformation and standardization
- New insights into the nature of biodiversity

This framework enables meta-analyses, cross-study comparisons, and deeper understanding of diversity patterns that were previously impossible.

## References

1. Information Theory and Diversity Metrics (In preparation)
2. Mathematical Foundations of Universal Diversity Decomposition (Submitted)

## Session Info

```{r session-info}
sessionInfo()
```